Higher-Order unification for free!

### 2024-04-22 paper/main.tex
conclusions

\noindent
However, this encoding has two big limitations. First it is
not always feasible to adopt it for Coq due to the fact that the type system
of the ML is too limited to accommodate for that one the object language,
e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\
Second, the encoding of Coq terms provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples, they cannot be symbols of the ML.

In the literature we could find a related encoding of the Calculus of
Constructions (CC)~\cite{felty93lics}. The goal of that work is to exhibit
a logic program performing proof checking for CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the one of CC. The encoding is hence tailored
toward a different goal, for example it utilizes three relations to represent the
equational theory of CC, and that choice alone makes things harder for us.
Section 6 contains a discussion about the use of the
unification procedure of the ML in presence of non ground goals, but
the authors do not aim at exploiting it to the degree we want.\todo{sucks}

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.

\emph{The approach presented in this paper}
provides a third option that addresses all the concerns
mentioned earlier. It capitalizes on the benefit of not requiring to fully
implement
the unification algorithm of the object language. Instead, it employs the
unification capabilities of the meta language, facilitated by the various links
%we establish
to manage ``problematic'' subterms. As a result of this choice our encoding takes
advantage of indexing data structures and mode analysis for clause filtering.
It it worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable by the meta language logic programming engine.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search, and apply forms of static analysis for
the meta language, such as determinacy, to the object language.
\todo{mention preliminar experiment show that our prolog is 2x faster than the Coq one} 
Finally, our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment, and
it is not tightly coupled with Coq.

We considered mechanizing our results with Abella~\cite{gacek2008abella}, especially in the early phases.
Unfortunately we could not. One reason
is that eliminating the cut operator, although possible, has a high cost in
verbosity.
% he same for the \elpiIn{name}
%\todo{is it sill described?} predicate.
Another problem is that the logic of Abella does not let one quantify over
predicates and combinators like \elpiIn{forall} are used pervasively in our
code.
\todo{finish}

>>>>

\noindent
However, this encoding has two big limitations. First it is
not always feasible to adopt it for Coq due to the fact that the type system
of the ML is too limited to accommodate for that one the object language,
e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\
Second, the encoding of Coq terms provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples, they cannot be symbols of the meta language.

In the literature we could find a related encoding of the Calculus of
Constructions (CC)~\cite{felty93lics}. The goal of that work is to exhibit
a logic program performing proof checking for CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the one of CC. The encoding is hence tailored
toward a different goal, for example it utilizes three relations to represent the
equational theory of CC, and that choice alone makes things harder for us.
Section 6 contains a discussion about the use of the
unification procedure of the meta language in presence of non ground goals, but
the authors are not interested in exploiting a decidable fragment of
higher-order unification but are reather leaning towards an interactive
system where explicit
control is given the user.

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.\\
~\\
\emph{The approach presented in this paper}
%provides a third option that 
addresses all the concerns
mentioned earlier. It takes advantage of the
unification capabilities of the meta language
at the price of handling ``problematic'' sub terms on
the side.\\
As a result %of this choice 
our encoding takes
advantage of indexing data structures and
mode analysis for clause filtering.
It it worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable by the meta language logic programming engine.
Some preliminary benchmarks on the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
encouraging results: our type-class solver has some
overhead on small goal but is consistently faster than
the Coq one starting from goals made of XXX nodes.\\
\todo{D: how many?}
Our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search. Finally, our shallow encoding can
enable to apply forms of static analysis for
the meta language, such as determinacy, also to the object language.

We considered mechanizing our results with Abella~\cite{gacek2008abella},
especially in the early phase of this work, but unfortunately we could not.
The main reason is that the logic of Abella does not let one quantify over
predicates nor user the cut operator. Eliminating cut or
combinators like \elpiIn{map} is surely possible, but 
has a high cost in verbosity given how pervasively they are used.
Moreover, and more importantly, it creates a distance
between the verified and the actual code, which we found to
hinder our exploratory phase.
